---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ASTR (Archaemetric Standards and Tools in R)

<!-- badges: start -->
[![R-CMD-check](https://github.com/archaeothommy/ASTR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/archaeothommy/ASTR/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/ASTR)](https://CRAN.R-project.org/package=ASTR)
[![Codecov test coverage](https://codecov.io/gh/archaeothommy/ASTR/graph/badge.svg)](https://codecov.io/gh/archaeothommy/ASTR)

<!-- badges: end -->

ASTR defines and implements a community reporting standard for archaeometric datasets. It also provides easy-to-use functions for common plots, statistical analyses, data processing, and data  transformation workflows in archaeometry. 

## Overview 

ASTR aims to increase reproducibility of data processing by facilitating the use of scripting languages in archeometry in the narrower sense, i.e., the material scientific investigation of usually inorganic archaeological materials. 

### Community reporting standard

The community reporting standard defines a [set of conventions on the structure and naming of datasets](vignettes/VG.ASTRschema.0.0.2.Rmd) to make them interoperable and allow automated handling of the data. Based on these conventions, ASTR

* Recognises and classifies isotope and chemical data, their units, and analytical precision, 
* Handles unit conversion on the fly, including non-SI units such as *ppm*, *at%*, and *cps* (counts per second). 

### Tool collection

The functions do not expect datasets according to the community reporting standard but their default values for their input  follow whereever possible its conventions. The collection provides easy-to-use functions for a wide range of tasks such as 

* Standard plots as [ggplot2](https://ggplot2.tidyverse.org/) geoms (e.g., spidergrams, KDE)
* Material classification (e.g., copper types)
* Data transformation and processing (e.g., calculation of $\delta$-values from standard-sample-bracketing measurements)
* Data conversion (e.g., calculation of Lead isotope age model parameters)
* Statistics (e.g, Distribution of data in a pointcloud of reference data)
* Normalisation of data to standard compositions (e.g., chondrite) 
* Unit conversion (e.g., at% to wt% and vice versa) 

See the full list on the [package website](https://archaeothommy.github.io/ASTR/index.html).

<!--
### Example

include here a small example for inputting a example data set and plotting it such as 

library(ASTR)
df <- import()

library(ggplot2)
ggplot(df) + 
 geom_spidergram()
-->

## Installation

You can install the development version of ASTR from [GitHub](https://github.com/) with:

``` {r, eval = FALSE}
 install.packages("pak")
 pak::pak("archaeothommy/ASTR")
```

## Getting started

We recommend reading the following resources to get familiar with the package: 

* Community reporting standard: [Conventions](vignettes/VG.ASTRschema.0.0.2.Rmd) and their  [explanation](vignettes/VG.ASTR.Conventions.explained.Rmd)

## Contributing and Code of Conduct

Please read our [contributor guide](https://archaeothommy.github.io/ASTR/CONTRIBUTING.html) to learn how you can contribute to ASTR. The ASTR project is released with a [Contributor Code of Conduct](https://archaeothommy.github.io/ASTR/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.  

## Acknowledgements and funding

Development was initiated with the workshop *Towards an Archaeological Science Toolbox in R “ASTR”* at the [Lorentz Center](https://www.lorentzcenter.nl/) in Leiden (The Netherlands). In addition to the in-kind funding of the Lorentz Center, the workshop received funding from the Stichting Nederlands Museum voor Anthropologie en Praehistorie (Foundation for Anthropology and Prehistory in the Netherlands). The workshop and further development of ASTR received funding from the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) – project 571205551. 
