---
title: "Working with ASTR objects"
csl: apa.csl
link-citations: TRUE
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with ASTR objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r data-prep, include=FALSE}
set.seed(123)
data <- data.frame(
  "Group" = c(rep("A", 5), rep("B", 5)),
  "Cu_wt%" = round(runif(10, 85, 95), 1),
  "Sn_wt%" = round(runif(10, 5, 10), 1),
  "As_wt%" = round(runif(10, 1.5, 2.5), 2),
  "Sb_wt%" = round(runif(10, 0.5, 1), 2),
  "Ag_ppm" = round(runif(10, 100, 350), 0),
  "Ni_ppm" = round(runif(10, 300, 650), 0),
  "d65Cu" = c(round(runif(6, -0.5, 0.5), 2), rep("#REF!", 4)),
  "206Pb/204Pb" = round(c(rep(NA, 5), runif(5, 18.3, 18.8)), 4),
  "206Pb/204Pb_err2SD" = round(c(rep(NA, 5), runif(5, 0.0007, 0.0011)), 5),
  "207Pb/204Pb" = round(c(rep(NA, 5), runif(5, 15.6, 15.7)), 4),
  "207Pb/204Pb_err2SD" = round(c(rep(NA, 5), runif(5, 0.0006, 0.0009)), 5),
  "208Pb/204Pb" = round(c(rep(NA, 5), runif(5, 38.0, 38.5)), 3),
  "208Pb/204Pb_err2SD" = round(c(rep(NA, 5), runif(5, 0.001, 0.002)), 4),
  check.names = FALSE
)
```

To make the most out of the features offered by the ASTR package, input data should be imported into `ASTR` objects. `ASTR` objects represent archaeometric and geochemical data in R in a standardized and semantically rich format, to allow for convenient and safe data analysis. If the input data structure adheres to the conventions set out in the [ASTR schema](VG.ASTRschema.0.0.2.html), then the package can automatically build valid `ASTR` objects from it. When working with the package, many of its functions require only an `ASTR` object, making them convenient to use even for users who consider themselves less experienced with R. This article highlights some of these functionalities.

Before we start, we want to load the ASTR package: 

```{r setup}
library(ASTR)
```

## Data import and inspection
To work with an `ASTR` object, we have to create one. The function `read_ASTR()` imports spreadsheets in a variety of formats, including Excel files. For this article, I created some mock-up data: 

```{r echo=FALSE}
data
```


Note how the column headers are organised according to the ASTR conventions. This allows the package to _understand_ some of the semantics of the dataset in the `read_ASTR()` process, and correctly represent them in the resulting `ASTR` object. A closer look at the mock-up data also reveals (common) problems with the copper isotope data: it seems that some Excel formulas did not work as intended and left `#REF!` entries. `read_ASTR()` will read these as `NA` values. Finally, every `ASTR` object requires one column that acts as a unique row identifier. In this dataset there is no such column, though, only a `Group` identifier. `read_ASTR()` can automatically turn such a column to a unique identifier.

We can now perform the reading process. As we are working with mock-up data in an R vignette, we do not read from the file system. We only need to call an essential internal function of `read_ASTR()`: `as_ASTR()`. It turns R data.frames to `ASTR` objects. If we would start from an Excel file we would call `read_ASTR()` instead.

To do this in practice we not only have to submit our `data` to `as_ASTR()`, but also set two other arguments: 1. We have to define one of the columns as the ID column with `id_column`, and 2. we explicitly have to mark any column providing contextual information with `context` (i.e. no analytical values, cf. [ASTR schema: Implementation](VG.ASTR.Schema.Implementation.html)).

```{r echo=TRUE}
data <- as_ASTR(data, id_column = "Group", context = c("Group"))
```

While ASTR attempts to make data import as easy as possible, it also aims to be transparent about it. Consequently, `as_ASTR()` issues several warnings when confronted with our mock-up data, flagging (potential) issues:

* The column for the ID did contain non-unique values but ASTR automatically created unique values from it.
* Some values in the column `d65Cu` were replaced with `NA`, when inspecting the data set, it is clear that these were the Excel `#REF!` error messages. In general, ASTR replaces by default common and user-defined indicators for missing values and also for indicators of "below detection limit" during import. 
* Across the dataset, 34 values are missing.

To learn more about the (potential) issues with our dataset, it suggests to run `validate()`: 

```{r echo=TRUE}
validate(data)
```

This summary of where to find the missing values comes in handy for large datasets! 

In our case, the dataset is fine. We might want to check on the missing values in the copper isotope data because apparently there was something wrong with the Excel file. In case of the lead isotope data, however, they are intentional (they were not measured for these samples).

Having a closer look on the imported data set shows, that ASTR did one more thing: 

```{r echo=TRUE}
data
```

Our relative unit "ppm" was converted into SI-units and the values now have a measurement unit R can understand (note that the unit `wt%` is written as `wtP`)^[The %-sign can not be used in custom units]: 

``` {r}
units(data$Ag[1])
```

When looking into the data structure, we also see that it also labelled the different columns as e.g. concentration or isotope ratio: 

``` {r}
str(data)
```

## Subsetting data

With this much information about our data, it is easy to subset data. Let's say that, for example, we want to check if our data can be differentiated into groups based on the chemical composition by creating biplots of all possible combinations. The `get_ ... _columns()`-family comes in very handy here: 

```{r}
pairs(get_concentration_columns(data)[-1])
```

Note that we have to exclude the first column here, because the first column of ASTR objects and their subsets is always the ID column to ensure that data can be clearly assigned to a single sample. Actually, we might only be interested in the main elements of the metal. In this case, we can take advantage that these are the only ones values in `wtP`: 

```{r}
pairs(get_unit_columns(data, units = "wtP")[-1])
```

### Unit conversions 

While continuing investigating the metal, we may want to estimate its approximate melting temperature and remember that we can get this information from a phase diagram. While our values for copper and tin are in `wt%`, the phase diagram in front of us is in `at%`. Converting from `wt%` to `at%` is probably faster than searching for a version of the phase diagram in `wt%`^[While ASTR fully supports use the native pipe operator `|>`, it maintains backward compatibility to older R versions by using `%>%`]:  

```{r}
library(magrittr) # import of pipe operator `%>%`

unify_concentration_unit(data, unit = "wtP") %>%
  wt_to_at()
```

We use a two-step approach here to be more accurate: The actual conversion from `wt%` to `at%` is done by `wt_to_at()`. However, because Ag and Ni are not in `wt%`, they need to be converted first. That's why we have to run `unify_concentration_unit()` first: is detects values with units that can be converted into the target unit and converts them all into the same unit. While it is nice to be scientifically accurate, we can get only a rough estimate anyway because of the significant amounts of arsenic and antimony in the metal. Therefore, converting only copper and tin would be sufficient: 

```{r}
wt_to_at(data, elements = c("Cu", "Sn"))
```

<!-- ### Geochemical normalisation
-->

## Working with data 

Another task in our investigation of the copper items is to identify the type of copper alloy our metal is made of. A first rough approach could be to use the classification suggested by @Bray2015. Not surprising, we see that they all belong to the group "As+Sb": 

```{r}
copper_group <- copper_group_bray(data)
copper_group
```

After we look into the chemical compositions, it is time to turn to the lead isotope data. Before plotting them, we want to calculate their age model parameters because they give us some idea about the geological setting. The model by @stacey1975 is still the most commonly used one. Because we have lead isotope data only for group B, we exclude data from group A: 

```{r warning=FALSE}
data_pb_iso <- data[data$Group == "B", ]
age_model_params <- pb_iso_age_model(data_pb_iso, model = "SK75")
age_model_params
```
<!--
## Visualisation

-->

## Data export

After finishing our analysis of the data, we want to save the copper type and lead isotope model age parameters to our hard drive. For this task, ASTR relies on the functions already provided by R and other R packages (e.g. `readr::write_csv()`). Before export, we have to make sure that the information about the units of the values is preserved by including it back in the column name: 

```{r}
data_unitless <- remove_units(data, recover_unit_names = TRUE)
data_unitless
```
## References
