---
title: "Binary plot with kernel density estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"Binary plot with kernel density estimation"} 
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Examples
The following examples showcase the different visualisation and density estimate calculations for the geom_kde2d function. 

## Basic biplot 
library(readr)
ad <- read_csv("data-raw/ArgentinaDatabase.csv")
View(ad)

ggplot(ad, 
  aes(x = `206Pb/204Pb`, 
  y = `207Pb/204Pb`, 
  fill = `Mining site`)) + 
  geom_kde2d()
#'
#' # Adjusting quantiles to show deciles
#' ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, fill = Species)) +
#'   geom_kde2d(quantiles = 10, alpha = 0.5)
#'
#' # Using min_prob to show only density regions above the median
#' ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, fill = Species)) +
#'   geom_kde2d(quantiles = 10, min_prob = 0.5)
#'
#' # Creating an outline effect
#' ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
#'   geom_kde2d(quantiles = 1, min_prob = 0, fill = NA)
#'
#'  # Creating an outline effect, and using coord_cartesian to expand the
#'  # plot area so the full KDE regions show without clipping
#' ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
#'   geom_kde2d(quantiles = 1, min_prob = 0, fill = NA) +
#'   coord_cartesian(xlim = c(4, 8.2), ylim = c(2, 4.5), clip = "off")
#'
#' # Example of fallback behavior
#' # Create a dataset where one group has too few points for density estimation
#' set.seed(123)
#' df <- data.frame(
#'   x = c(rnorm(50), rnorm(50, 5), rnorm(2, 10)),
#'   y = c(rnorm(50), rnorm(50, 5), rnorm(2, 10)),
#'   group = rep(c("A", "B", "C"), c(50, 50, 2))
#' )
#'
#' # A message will indicate that group "C" is plotted as points
#' ggplot(df, aes(x, y, fill = group, colour = group)) +
#'   geom_kde2d(alpha = 0.4) +
#'   theme_minimal()
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ASTR)
```
