---
title: "Binary plot with kernel density estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Binary plot with kernel density estimation} 
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This vignette introduces the "Kde2" function which generates a binary plot with kernel density estimation.

Kernel density estimation (KDE) is a non-parametric method to transform continuous data into a smoothed probability density function. According to Hsu (2024) and Hsu et al. (2018) KDE has become popular in archaeological science publications mainly due to three advantages when analysing data:

1. They do not assume the normality of the data;
2. They can produce smoother distributions than conventional histograms, whose appearance is significantly affected by the choices of bin width and the start/end points of bins; and
3. They can represent data in a multidimensional space and enable users to effectively compare different datasets either graphically or mathematically (see De Ceuster et al. 2025 also for the advantages of the use of one-dimensional KDE’s).

## What is Kernel density estimation (KDE)?







# Examples
The following examples showcase the different visualisation and density estimate calculations for the geom_kde2d function. 

## Basic biplot 
library(readr)
ad <- read_csv("data-raw/ArgentinaDatabase.csv")
View(ad)

ggplot(ad, 
  aes(x = `206Pb/204Pb`, 
  y = `207Pb/204Pb`, 
  fill = `Mining site`)) + 
  geom_kde2d()
#'
#' # Adjusting quantiles to show deciles
#' ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, fill = Species)) +
#'   geom_kde2d(quantiles = 10, alpha = 0.5)
#'
#' # Using min_prob to show only density regions above the median
#' ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, fill = Species)) +
#'   geom_kde2d(quantiles = 10, min_prob = 0.5)
#'
#' # Creating an outline effect
#' ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
#'   geom_kde2d(quantiles = 1, min_prob = 0, fill = NA)
#'
#'  # Creating an outline effect, and using coord_cartesian to expand the
#'  # plot area so the full KDE regions show without clipping
#' ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
#'   geom_kde2d(quantiles = 1, min_prob = 0, fill = NA) +
#'   coord_cartesian(xlim = c(4, 8.2), ylim = c(2, 4.5), clip = "off")
#'
#' # Example of fallback behavior
#' # Create a dataset where one group has too few points for density estimation
#' set.seed(123)
#' df <- data.frame(
#'   x = c(rnorm(50), rnorm(50, 5), rnorm(2, 10)),
#'   y = c(rnorm(50), rnorm(50, 5), rnorm(2, 10)),
#'   group = rep(c("A", "B", "C"), c(50, 50, 2))
#' )
#'
#' # A message will indicate that group "C" is plotted as points
#' ggplot(df, aes(x, y, fill = group, colour = group)) +
#'   geom_kde2d(alpha = 0.4) +
#'   theme_minimal()


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ASTR)
```

# Introduction

This vignette introduces the `geom_kde2d` function which generates a two-dimensional plot with kernel density estimation.

Kernel density estimation (KDE) is a non-parametric method to transform continuous data into a smoothed probability density function. According to Hsu (2024,Hsu et al. 2018) KDE has become popular in archeaological science publications mainly due to three advantages when analysing data:

They do not assume the normality of the data; They can produce smoother distributions than conventional histograms, whose appearance is significantly affected by the choices of bin width and the start/end points of bins; and They can represent data in a multidimensional space and enable users to effectively compare different datasets either graphically or mathematically (see De Ceuster et al. 2025 also for the advantages of the use of one-dimensional KDE’s).

# Function workflow

The geom_kde2 function calculates the 2D kernel density estimate using the `kde` function from the ks package (Duong 2007; 2025), while handling the contours manually (compute.cont = FALSE).

It then calculates the actual density values (the contour levels) corresponding to the probability levels (probs) using the `ks::contourLevels` function.

Using the base R function `grDevices::contourLines`, the contour line coordinates are generated, using the grid points (eval.points) and the density estimate (estimate) from the kde object, and the calculated levels. The contour coordinates are converted into a data frame as list items. The function extracts its coordinates (x, y) and assigns a unique group ID for each contour. Specific error messages are printed for failed density calculations.

The original data are then returned with the added type column and are checked for fallback (the type column equals "points") to determine if the density estimates failed). A data frame for plotting points using the original x and y data from the fallback data is created, applying the determined aesthetics and these are drawn using the in-built GeomPoint `draw_panel`.

Default aesthetic values are applied if unchanged by the user.

# Examples

The following examples showcase the different visualisation and density estimate calculations for the geom_kde2d function.

## Basic biplot

```{r}
# load example Lead isotope data that is included with the package
data(ArgentinaDatabase)

library(ggplot2)

ggplot(ArgentinaDatabase, 
  aes(x = `206Pb/204Pb`, 
      y = `207Pb/204Pb`, 
  fill = `Mining site`)) + 
  geom_kde2d()
```

This example reads in the sample dataset and calculates the kernel density estimates in the function `geom_kde2d()`, colouring by the `Mining site` variable from the dataset. Warning messages appear for each group where the sample size is too small to calculate the estimates: `No density estimate possible for group 'x' plotting points instead. The leading minor of order 1 is not positive`. In those cases, the plotting falls back to plotting individual points from the original dataset.

## Biplot with adjusted quantiles to show deciles

The default function reverts to 4 quantiles. This examples illustrates the quantiles set to 10 to show deciles. It also sets the transparency to 0.5 (`alpha`).

```{r}
ggplot(ArgentinaDatabase, 
  aes(x = `206Pb/204Pb`, 
      y = `207Pb/204Pb`, 
      fill = `Mining site`)) + 
  geom_kde2d(quantiles = 10, 
             alpha = 0.5)
```

## Biplot with the minimum probability argument added to show only density regions above the median

```{r}
ggplot(ArgentinaDatabase, 
  aes(x = `206Pb/204Pb`, 
      y = `207Pb/204Pb`, 
      fill = `Mining site`)) + 
  geom_kde2d(quantiles = 10, 
             min_prob = 0.5)
```

## Creation of an outline effect around the density region

For this effect the fill argument is set to `NA`

```{r}
ggplot(ArgentinaDatabase, 
  aes(x = `206Pb/204Pb`, 
      y = `207Pb/204Pb`, 
      colour = `Mining site`)) + 
  geom_kde2d(quantiles = 1, 
             min_prob = 0, 
             fill = NA)
```

## Creation of an outline effect around the density region solving clipping issues

In some cases density regions can be clipped by the plot area. The addition of the `coord_cartesia`n function from `ggplot2` expands the plot area so the full density regions are shown without clipping.The limits are set for the axes using the `xlim` and `ylim` arguments.

```{r}
ggplot(ArgentinaDatabase, 
  aes(x = `206Pb/204Pb`, 
      y = `207Pb/204Pb`, 
      colour = `Mining site`)) + 
  geom_kde2d(quantiles = 1,
             min_prob = 0, 
             fill = NA) + 
  coord_cartesian(xlim = c(4, 8.2), 
                  ylim = c(2, 4.5), 
                  clip = "off")
```
