---
title: "Binary plot with kernel density estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Binary plot with kernel density estimation} 
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This vignette introduces the "Kde2" function which generates a binary plot with kernel density estimation.

Kernel density estimation (KDE) is a non-parametric method to transform continuous data into a smoothed probability density function. According to Hsu (2024) and Hsu et al. (2018) KDE has become popular in archaeological science publications mainly due to three advantages when analysing data:

1. They do not assume the normality of the data;
2. They can produce smoother distributions than conventional histograms, whose appearance is significantly affected by the choices of bin width and the start/end points of bins; and
3. They can represent data in a multidimensional space and enable users to effectively compare different datasets either graphically or mathematically (see De Ceuster et al. 2025 also for the advantages of the use of one-dimensional KDE’s).

## What is Kernel density estimation (KDE)?

Kernel density estimation is a widely-used tool for visualising the distribution of data.

Probability density function (pdf) shows how the whole 100% probability mass is distributed over the x-axis, i.e., over the values of an X random variable. Kernel estimation of probability density function produces a smooth empirical pdf based on individual locations of all sample data. Such pdf estimate seems to better represent the "true" pdf of a continuous variable (Węglarczyk 2018). 

###Kernel density estimation calculation

Let the series \( \{x_1, x_2, \ldots, x_n\} \) be an independent and identically distributed (i.i.d.) sample of \( n \) observations taken from a population \( X \) with an unknown probability distribution function \( f(x) \).

The kernel estimate \( \hat{f}(x) \) of the original \( f(x) \) assigns each \( i \)-th sample data point \( x_i \) a function \( K(x_i, t) \), called a *kernel function*, in the following way [11]:

\[
\hat{f}(t) = \frac{1}{n} \sum_{i=1}^{n} K(x_i, t)
\tag{1}
\]

The kernel function \( K(x, t) \) is nonnegative and bounded for all \( x \) and \( t \):

\[
0 \leq K(x, t) < \infty \quad \text{for all real } x, t
\tag{2}
\]

and, for all real \( x \),

\[
\int_{-\infty}^{\infty} K(x, t) \, dt = 1.
\tag{3}
\]

Property (3) ensures the required normalization of the kernel density estimate given in (1).

\[
\int_{-\infty}^{\infty} \hat{f}(t) \, dt 
= \frac{1}{n} \sum_{i=1}^{n} \int_{-\infty}^{\infty} K(x_i, t) \, dt = 1.
\tag{4}
\]

In other words, the kernel transforms the *"sharp"* (point) location of \( x_i \) into an interval centred (symmetrically or not) around \( x_i \).

(PDF) Węglarczyk, S.2018. Kernel density estimation and its application. Available from: https://www.researchgate.net/publication/328785939_Kernel_density_estimation_and_its_application [accessed Oct 24 2025].

(PDF) Kernel density estimation and its application. Available from: https://www.researchgate.net/publication/328785939_Kernel_density_estimation_and_its_application [accessed Oct 24 2025].


###Further reading
Simonoff, J.S. (1996). *Smoothing Methods in Statistics*. Springer-Verlag, New York.


# Examples
The following examples showcase the different visualisation and density estimate calculations for the geom_kde2d function. 

## Basic biplot 
library(readr)
ad <- read_csv("data-raw/ArgentinaDatabase.csv")
View(ad)

ggplot(ad, 
  aes(x = `206Pb/204Pb`, 
  y = `207Pb/204Pb`, 
  fill = `Mining site`)) + 
  geom_kde2d()
#'
#' # Adjusting quantiles to show deciles
#' ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, fill = Species)) +
#'   geom_kde2d(quantiles = 10, alpha = 0.5)
#'
#' # Using min_prob to show only density regions above the median
#' ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, fill = Species)) +
#'   geom_kde2d(quantiles = 10, min_prob = 0.5)
#'
#' # Creating an outline effect
#' ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
#'   geom_kde2d(quantiles = 1, min_prob = 0, fill = NA)
#'
#'  # Creating an outline effect, and using coord_cartesian to expand the
#'  # plot area so the full KDE regions show without clipping
#' ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
#'   geom_kde2d(quantiles = 1, min_prob = 0, fill = NA) +
#'   coord_cartesian(xlim = c(4, 8.2), ylim = c(2, 4.5), clip = "off")
#'
#' # Example of fallback behavior
#' # Create a dataset where one group has too few points for density estimation
#' set.seed(123)
#' df <- data.frame(
#'   x = c(rnorm(50), rnorm(50, 5), rnorm(2, 10)),
#'   y = c(rnorm(50), rnorm(50, 5), rnorm(2, 10)),
#'   group = rep(c("A", "B", "C"), c(50, 50, 2))
#' )
#'
#' # A message will indicate that group "C" is plotted as points
#' ggplot(df, aes(x, y, fill = group, colour = group)) +
#'   geom_kde2d(alpha = 0.4) +
#'   theme_minimal()


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ASTR)
```

# Introduction

This vignette introduces the `geom_kde2d` function which generates a binary plot with kernel density estimation.

Kernel density estimation (KDE) is a non-parametric method to transform continuous data into a smoothed probability density function. According to Hsu (2024,Hsu et al. 2018) KDE has become popular in archeaological science publications mainly due to three advantages when analysing data:

They do not assume the normality of the data;
They can produce smoother distributions than conventional histograms, whose appearance is significantly affected by the choices of bin width and the start/end points of bins; and
They can represent data in a multidimensional space and enable users to effectively compare different datasets either graphically or mathematically (see De Ceuster et al. 2025 also for the advantages of the use of one-dimensional KDE’s).

# Examples
The following examples showcase the different visualisation and density estimate calculations for the geom_kde2d function. 

## Basic biplot 

```{r}
# load example Lead isotope data that is included with the package
data(ArgentinaDatabase)

ggplot(ArgentinaDatabase, 
  aes(x = `206Pb/204Pb`, 
  y = `207Pb/204Pb`, 
  fill = `Mining site`)) + 
  geom_kde2d()
```
  
This example reads in the sample dataset and calculates the kernel density estimates in the function geom_kde2d(), colouring by the Mining site variable from the dataset. Warning messages appear for each group where the sample size is too small to calculate the estimates: No density estimate possible for group 'x' plotting points instead.
the leading minor of order 1 is not positive. In those cases, the plotting falls back to plotting individual points from the original dataset.

## Biplot with adjusted quantiles to show deciles
The default function reverts to 4 quantiles. This examples illustrates the quantiles set to 10 to show deciles.It also sets the transparency to 0.5. 

```{r}
ggplot(ArgentinaDatabase, 
  aes(x = `206Pb/204Pb`, 
  y = `207Pb/204Pb`, 
  fill = `Mining site`)) + 
  geom_kde2d(quantiles = 10, alpha = 0.5)
```
  
## Biplot with the minimum probability argument added to show only density regions above the median

```{r}
ggplot(ArgentinaDatabase, 
  aes(x = `206Pb/204Pb`, 
  y = `207Pb/204Pb`, 
  fill = `Mining site`)) + 
  geom_kde2d(quantiles = 10, min_prob = 0.5)
```
  
## Creation of an outline effect around the density region
For this effect the fill argument is set to NA

```{r}
ggplot(ArgentinaDatabase, 
  aes(x = `206Pb/204Pb`, 
  y = `207Pb/204Pb`, 
  colour = `Mining site`)) + 
  geom_kde2d(quantiles = 1, min_prob = 0, fill = NA)
```

##   Creation of an outline effect around the density region solving clipping issues
In some cases density regions can be clipped by the plot area. The addition of the `coord_cartesia`n function from ggplot2 expands the plot area so the full density regions are shown without clipping.The limits are set for the axes using the xlim and ylim arguments. 

```{r}
ggplot(ArgentinaDatabase, 
  aes(x = `206Pb/204Pb`, 
  y = `207Pb/204Pb`, 
  colour = `Mining site`)) + 
  geom_kde2d(quantiles = 1, min_prob = 0, fill = NA) + 
  coord_cartesian(xlim = c(4, 8.2), ylim = c(2, 4.5), clip = "off")
```


